---
title: "Bike rental"
author: "Ziling Yang"
date: "9/11/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(GGally)
```

```{r read file}
bikedata <- read.csv("day.csv",header=T)
names(bikedata)
```

# How temperature affects bike count


```{r transform data}
#Transform temp and atemp to o.temp and o.atemp without division
bikedata <- 
  bikedata %>% mutate(actual.temp = temp*41) %>% 
  mutate(actual.atemp = atemp*50) %>%
  mutate(actual.windspeed = windspeed*67) %>%
  mutate(actual.hum = hum*100)
  

```


# Regressing bike count with actual temperature
```{r simple linear model for actual temperature against bike count}
#Perform a simple linear regression 
actual_temp.simple.regress = (lm(bikedata$cnt ~ bikedata$actual.temp))
summary(actual_temp.simple.regress)

```


```{r scatter for temperature against rental count}
ggplot(bikedata, aes(actual.temp, cnt)) + geom_point(shape=1) + geom_abline(intercept = coef(actual_temp.simple.regress)[1], slope = coef(actual_temp.simple.regress)[2], colour = "red") + ylab('Total bike rental') + xlab('actual temperature') + labs(title = 'temperature against rental count (model imposed)')
```

```{r plot }
plot(actual_temp.simple.regress)
```
# How temperature affects casual and registered bikers

```{r how temperature affects casual bikers}

casual_temp.regress = (lm(bikedata$casual ~ bikedata$actual.temp))
summary(casual_temp.regress)

```

```{r how temperature affects registered bikers}

registered_temp.regress = (lm(bikedata$registered ~ bikedata$actual.temp))
summary(registered_temp.regress)

```




```{r}

ggplot(bikedata, aes(x = actual.temp)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) + geom_abline(intercept = coef(registered_temp.regress)[1], slope = coef(registered_temp.regress)[2], colour = "blue") +  geom_abline(intercept = coef(casual_temp.regress)[1], slope = coef(casual_temp.regress)[2], colour = "red") + ylab('bike rental count') + xlab('actual temperature') + labs(title = 'temperature against rental count (model imposed)', labels=c("registered", "casual"))
```


# Feel temperature and bike rental count
```{r simple linear model for feel temperature against bike count}
#Perform a simple linear regression 
feeltemp.simple.regress = (lm(bikedata$cnt ~ bikedata$actual.atemp))
summary(feeltemp.simple.regress)

```

```{r scatter for feel temp against rental count}
ggplot(bikedata, aes(actual.temp, cnt)) + geom_point(shape=1) + geom_abline(intercept = coef(feeltemp.simple.regress)[1], slope = coef(feeltemp.simple.regress)[2], colour = "red") + ylab('Total bike rental') + xlab('feel temperature') + labs(title = 'feel temperature against rental count (model imposed)')
```
# How feel temperature affects casual and registered bikers


```{r how feel temperature affects casual bikers}

casual_feeltemp.regress = (lm(bikedata$casual ~ bikedata$actual.atemp))
summary(casual_feeltemp.regress)

```


```{r how feel temperature affects registered bikers}

registered_feeltemp.regress = (lm(bikedata$registered ~ bikedata$actual.atemp))
summary(registered_feeltemp.regress)

```

```{r}

ggplot(bikedata, aes(x = actual.atemp)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) + geom_abline(intercept = coef(registered_feeltemp.regress)[1], slope = coef(registered_feeltemp.regress)[2], colour = "blue") +  geom_abline(intercept = coef(casual_feeltemp.regress)[1], slope = coef(casual_feeltemp.regress)[2], colour = "red") + ylab('bike rental count') + xlab('feel temperature') + labs(title = 'feel temperature against rental count (model imposed)', labels=c("registered", "casual")) 
```


# Correlation between variables

```{r correlation matrix}
cor(bikedata$actual.atemp, bikedata$cnt)
cor(bikedata$actual.atemp, bikedata$casual)
cor(bikedata$actual.atemp, bikedata$registered)
cor(bikedata$actual.temp, bikedata$cnt)
cor(bikedata$actual.temp, bikedata$casual)
cor(bikedata$actual.temp, bikedata$registered)
cor(bikedata$actual.atemp, bikedata$casual)
cor(bikedata$actual.atemp, bikedata$actual.temp)

```

# Multiple linear Regression

Use covariates holiday, weekday, weathersit, temp, hum ,windspeed to regress total rental bike counts. 

First we regress with temperature^2. 

```{r higher power regression with temperature}
m.quadls_casual <- lm(bikedata$casual ~ bikedata$actual.temp + I(bikedata$actual.temp^2))
m.quadls_registered <- lm(bikedata$registered ~ bikedata$actual.temp + I(bikedata$actual.temp^2))
```


```{r}
ggplot(bikedata, aes(x = actual.temp)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) +   geom_line(data = fortify(m.quadls_casual), aes(x = bikedata$actual.temp, y = .fitted), color = "red") + geom_line(data = fortify(m.quadls_registered), aes(x = bikedata$actual.temp, y = .fitted), color = "blue") + labs(title = "Scatter plot with fitted quadratic model")
```
```{r}
StanResQLS_casual <- rstandard(m.quadls_casual)
ggplot(bikedata, aes(x = actual.temp, y=StanResQLS_casual)) + geom_point(shape = 1) + geom_hline(yintercept = -2) + geom_hline(yintercept = 2) + ggtitle("Standarized Residuals")
```
```{r}
StanResQLS_registered <- rstandard(m.quadls_registered)
ggplot(bikedata, aes(x = actual.temp, y=StanResQLS_registered)) + geom_point(shape = 1) + geom_hline(yintercept = -2) + geom_hline(yintercept = 2)+ ggtitle("Standarized Residuals")
```

Most data points are within two standard deviations within the prediction. This means that the model is a good fit. 

```{r}
summary(m.quadls_casual)
```

```{r}
summary(m.quadls_registered)
```

```{r}
bikedata$season <- as.factor(bikedata$season)
bikedata$holiday <- as.factor(bikedata$holiday)
bikedata$weathersit <- as.factor(bikedata$weathersit)
bikedata$workingday <- as.factor(bikedata$workingday)
bikedata$yr <- as.factor(bikedata$yr)
```

```{r}
m.mls_casual <- lm(bikedata$casual ~ bikedata$holiday + bikedata$weathersit + bikedata$workingday +  bikedata$actual.hum + bikedata$actual.temp+ bikedata$actual.windspeed)

summary(m.mls_casual)
```
```{r}
m.mls_registered <- lm(bikedata$registered ~ bikedata$holiday + bikedata$weathersit + bikedata$workingday + bikedata$yr + bikedata$actual.hum + bikedata$actual.temp+ bikedata$actual.windspeed)

summary(m.mls_registered)
```

```{r}
data <- data.frame(bikedata$casual, bikedata$holiday, bikedata$weathersit, bikedata$workingday, bikedata$actual.hum, bikedata$actual.temp, bikedata$actual.windspeed)

ggpairs(data, lower = list(continuous = wrap("points", alpha = 0.3, size= 0.1)))
```

```{r}
data <- data.frame(bikedata$registered, bikedata$holiday, bikedata$weathersit, bikedata$workingday, bikedata$actual.hum, bikedata$actual.temp, bikedata$actual.windspeed)

ggpairs(data, lower = list(continuous = wrap("points", alpha = 0.3, size= 0.1)))
```
```{r}
StanResMLS_registered <- rstandard(m.mls_registered)
ggplot(bikedata, aes(x = actual.temp, y=StanResMLS_registered)) + geom_point(shape = 1) + geom_hline(yintercept = -2) + geom_hline(yintercept = 2)+ ggtitle("Standarized Residuals")
```

```{r}
StanResMLS_casual <- rstandard(m.mls_casual)
ggplot(bikedata, aes(x = actual.temp, y=StanResMLS_casual)) + geom_point(shape = 1) + geom_hline(yintercept = -2) + geom_hline(yintercept = 2)+ ggtitle("Standarized Residuals")
```

