---
title: "Deliverable 4"
output: github_document
---

# Preprocessing

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(GGally)
library(caret)
library(kableExtra)
```

```{r read file}
bikedata <- read.csv("day.csv",header=T)
names(bikedata)
```

```{r transform data}
#Transform temp, atemp, windspeed, and humidity to actual values
bikedata <- 
  bikedata %>% mutate(actual.temp = temp*41) %>% 
  mutate(actual.atemp = atemp*50) %>%
  mutate(actual.windspeed = windspeed*67) %>%
  mutate(actual.hum = hum*100)
#process factor data
bikedata$season <- as.factor(bikedata$season)
bikedata$holiday <- as.factor(bikedata$holiday)
bikedata$weathersit <- as.factor(bikedata$weathersit)
bikedata$workingday <- as.factor(bikedata$workingday)
bikedata$yr <- as.factor(bikedata$yr)

bikedata <- bikedata %>% mutate(weekend = if_else(weekday == 0|weekday==6,1,if_else(weekday ==1|weekday==2|weekday==3|weekday==4|weekday==5,0,NA_real_) ))



```



```{r}
bikedata$weekend <- as.factor(bikedata$weekend)
```



```{r seperate data}
#Seperate training and validating datas base on year

training_d = bikedata %>%  filter(yr == 0)
set.seed(42)
#partitiontraining <- createDataPartition(y = train$cnt, p = 0.8, list = F)
#training_d <- train[partitiontraining, ]
#test_d <- train[-partitiontraining, ]
validate_d <- bikedata %>% filter(yr == 1)
```

# Season

We suspect that there will be a different distribution for casual and registered count in different seasons, therefore we graph them seperately.

```{r}
ggplot(training_d,aes(x=season,y=casual))+geom_col()
```

```{r}
ggplot(training_d,aes(x=season,y=registered))+geom_col()
```

The graphs show that for both casual and registered bikers, there are the most rental counts during austumn season and the least during the spring season. However, for registered, there are about the same amount of count during summer and winter while for casual there are significantly less counts during winter than during summer. Therefore we think that we should fit different models for registered and casual.

# Holiday
```{r}
ggplot(training_d,aes(x=holiday,y=casual,fill=season))+geom_col()
```

```{r}
ggplot(training_d,aes(x=holiday,y=registered,fill=season))+geom_col()
```

# Working day

```{r}
ggplot(training_d,aes(x=workingday,y=casual,fill=season))+geom_col()
```


```{r}
ggplot(training_d,aes(x=workingday,y=registered,fill=season))+geom_col()
```

# Weekday

```{r}
ggplot(training_d,aes(x=weekday,y=casual,fill=season))+geom_col()
```


```{r}
ggplot(training_d,aes(x=weekday,y=registered,fill=season))+geom_col()
```

There seems to be a difference in bike counts for both registered and casual between weekends and not weekends so we generate a new one hot variable. 0 when it is weekday 1 to 5 and 1 if it is 0 or 6. 


# Weather situation 

```{r}
ggplot(training_d,aes(x=weathersit,y=casual,fill=season))+geom_col()
```

```{r}
ggplot(training_d,aes(x=weathersit,y=registered,fill=season))+geom_col()
```

# correlation of continuous variables with rental count

```{r}
data <- data.frame(training_d$casual, training_d$actual.hum, training_d$actual.temp, training_d$actual.windspeed, training_d$actual.atemp)

data = data%>% rename( casual = training_d.casual, humidity= training_d.actual.hum,  temperature= training_d.actual.temp, windspeed = training_d.actual.windspeed, feeltemp = training_d.actual.atemp)

ggpairs(data, lower = list(continuous = wrap("points", alpha = 0.3, size= 0.7)))
```

```{r}
data <- data.frame(training_d$registered, training_d$actual.hum, training_d$actual.temp, training_d$actual.windspeed, training_d$actual.atemp)

data = data%>% rename( registered = training_d.registered, humidity= training_d.actual.hum,  temperature= training_d.actual.temp, windspeed = training_d.actual.windspeed, feeltemp = training_d.actual.atemp)

ggpairs(data, lower = list(continuous = wrap("points", alpha = 0.3, size= 0.7)))
```

The scatter correlation suggests that feel temperature and actual temperature are highly correlated because the scatter plot of temperature against feel temperature resembles a straight linear line and the correlation between them is 0.996, which is extremely close to 1. Therefore we will exclude feel temperature from our model to avoid collinearity. 

Humidity and casual and registered counts are not correlated as suggested by the low correlation value of -0.032 and 0.019 respectively. The scatter plot of casual and registered counts against humidity also shows a slope of around 0. Therefore humidity will likely not explain much of the variation in rental counts and we will exclude it from our model. 

The scatter plot of causual and registered against temperature also suggests a non linear relationship between temperature and rental counts because the plots decreases above a certain temperature. Therefore we should add higher temperature powers. 



# If workday is encaptured by holiday and weekday

```{r}
workingdayC_byHoliday <- training_d %>% count(workingday, holiday)
workingdayC_byHoliday <- workingdayC_byHoliday %>% filter(workingday ==1)
ggplot(workingdayC_byHoliday) + geom_col(aes(x=holiday, y=n))
```


```{r}
workingdayC_byWeekday <- training_d %>% count(workingday, weekend)
workingdayC_byWeekday <- workingdayC_byWeekday %>% filter(workingday ==1)
ggplot(workingdayC_byWeekday) + geom_col(aes(x=weekend, y=n))
```

We will exclude holiday and weekday from out model because the bar plot of holiday against working day count shows that all working days are not a holiday therefore workingday already encaptured the information of holidays. We will also exclude weekday from our model because the count of workingday is 0 on saturday and sunday therefore workingday also has already encaptured the information of weekday. 

# correlation between season and temperature

```{r}
ggplot(training_d,aes(x=season,y=actual.temp))+geom_boxplot()
```

The box plot of season against temperature suggests that there is a correlation between seasons and temperature. Autumn has the highest temperature and spring has the lowest temperature. We might want to exclude season to avoid collinearity in our model. 

# temperature and rental counts

```{r higher power regression with temperature}
m.quadls_casual <- lm(training_d$casual ~ training_d$actual.temp + I(training_d$actual.temp^2))
m.quadls_registered <- lm(training_d$registered ~ training_d$actual.temp + I(training_d$actual.temp^2))
```


```{r}
ggplot(training_d, aes(x = actual.temp)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) +   geom_line(data = fortify(m.quadls_casual), aes(x = training_d$actual.temp, y = .fitted), color = "red") + geom_line(data = fortify(m.quadls_registered), aes(x = training_d$actual.temp, y = .fitted), color = "blue") + labs(title = "Scatter plot with fitted quadratic model")
```

Therefore we conclude that we should include a quadratic term on temperature. 

# Wind speed and rental counts

```{r}
m.lin_casual <- lm(training_d$casual ~ training_d$actual.windspeed)
m.lin_registered <- lm(training_d$registered ~ training_d$actual.windspeed)
```

```{r}
ggplot(training_d, aes(x = actual.windspeed)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) +   geom_line(data = fortify(m.lin_casual), aes(x = training_d$actual.windspeed, y = .fitted), color = "red") + geom_line(data = fortify(m.lin_registered), aes(x = training_d$actual.windspeed, y = .fitted), color = "blue") + labs(title = "Scatter plot with fitted quadratic model")
```

# Model
```{r}
model.casual <- lm(casual ~ actual.windspeed + actual.temp + I(actual.temp^2) + weathersit + workingday, data = training_d)

model.registered <- lm(registered ~ actual.windspeed + actual.temp + I(actual.temp^2) + workingday +  weathersit, data = training_d)
```


```{r}
summary(model.casual)
```

```{r}
summary(model.registered)
```

All of the p values on the coefficients of the regressors are less than 0.005. Therefore we are confident that all the regressors have an effect on the rental counts individually. Furthermore, the p value of the F-statistic is less than 0.005. Therefore we are very confident that all the regressors are jointly significant. The R^2 value is arount 0.7, so the models explain around 70 percent of the variation in rental counts. (explain more in paper).

# Model diagnosis

```{r}
StanRes.casual <- rstandard(model.casual)
StanRes.registered <- rstandard(model.registered)

```

```{r}
ggplot() +
geom_point(data=training_d, aes(x=casual, y=StanRes.casual, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for casual bikers")
```


```{r}
ggplot() +
geom_point(data=training_d, aes(x=registered, y=StanRes.registered, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for registered bikers")
```

```{r}
Fitted_registered = fitted(model.registered)
ggplot() +
geom_point(aes(x=Fitted_registered, y=StanRes.registered, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals MLS Plot (Fitted) for registered bikers")
```

```{r}
Fitted_casual = fitted(model.casual)
ResFittedMLS_casual <- ggplot() +
geom_point(aes(x=Fitted_casual, y=StanRes.casual, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals MLS Plot (Fitted) for casual bikers")

ResFittedMLS_casual
```

```{r}
p <- ggplot(data.frame(StanRes.casual), aes(sample = StanRes.casual)) +
ggtitle("QQ MLS Plot for casual bikers")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanRes.registered), aes(sample = StanRes.registered)) +
ggtitle("QQ MLS Plot for registered bikers")
p + stat_qq() + stat_qq_line()
```
The fitted residual plot and the residual plot suggest that there are extreme outliers in the casual model and that the residual for both models are not evenly distributed around 0, therefore suggesting that there exists heterogeneity in the models. 

The QQ plots show a line that is roughly straight, therefore we conclude that the data of registered bikers come from a normally distributed sample. We can also conclude the same for casual bikers, however, there exists some data points that do not come from a normal distribution as indicated by the few datapoints that deviate significantly from the straight line.

We believe that a wls model will fit better given the residual plots.

(elaborate in paper)

# Weighted model

```{r}
weights_casual <- 1 / lm(abs(model.casual$residuals) ~model.casual$fitted.values)$fitted.values^2

wls.casual <- lm(casual ~ actual.windspeed + actual.temp + I(actual.temp^2) + workingday +  weathersit, data = training_d, weights = weights_casual)
summary.lm(wls.casual)

```

```{r}
weights_registered <- 1 / lm(abs(model.registered$residuals) ~model.registered$fitted.values)$fitted.values^2

wls.registered <- lm(registered ~ actual.windspeed + actual.temp + I(actual.temp^2) + workingday +  weathersit, data = training_d, weights = weights_registered)
summary.lm(wls.registered)

```

The F statistics of both weighted least sqare models become larger, which means that the in the weighted models, the regressors are more jointly significant than the ols models. (elaborate more in paper about coeff, p values, r values).

# Diagnosis of weighted least squares model

```{r}
StanResWLS.casual <- rstandard(wls.casual)
StanResWLS.registered <- rstandard(wls.registered)
```

```{r}
ggplot() +
geom_point(data=training_d, aes(x=casual, y=StanResWLS.casual, color = "WLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("WLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals WLS Plot for casual bikers")
```


```{r}
ggplot() +
geom_point(data=training_d, aes(x=registered, y=StanResWLS.registered, color = "WLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("WLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals WLS Plot for registeredl bikers")
```

```{r}
FittedWLS_casual = fitted(wls.casual)
ggplot() +
geom_point(aes(x=FittedWLS_casual, y=StanResWLS.casual, color = "WLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_casual, y=StanRes.casual, color = "MLS"), size = 1) +
scale_color_manual(name = element_blank(), labels = c("WLS", "MLS"), values = c("blue", "red")) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals WLS Plot (Fitted) for casual bikers") 
```

```{r}
FittedWLS_registered = fitted(wls.registered)
ggplot() +
geom_point(aes(x=FittedWLS_registered, y=StanResWLS.registered, color = "WLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_registered, y=StanRes.registered, color = "MLS"), size = 1) +
scale_color_manual(name = element_blank(), labels = c("WLS", "MLS"), values = c("blue", "red")) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals WLS Plot (Fitted) for registered bikers")
```


```{r}
p <- ggplot(data.frame(StanResWLS.casual), aes(sample = StanResWLS.casual)) +
ggtitle("QQ WLS Plot for casual bikers")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanResWLS.registered), aes(sample = StanResWLS.registered)) +
ggtitle("QQ WLS Plot for registered bikers")
p + stat_qq() + stat_qq_line()
```


Residual plots show less extreme outliers and more homogeneity but still exists heterogeneity and outliers. (Ask Julio)

# Validation

```{r}
p.casual <- predict(wls.casual, validate_d)
error.casual <- (p.casual- validate_d$casual)
RMSE_validation.caual <- sqrt(mean(error.casual^2))
RMSEWLS.casual <- sqrt(mean(resid(wls.casual)^2))

```

Square root mean square error for validation data set
```{r}
RMSE_validation.caual
```
square root mean square error for training data set
```{r}
RMSEWLS.casual
```

```{r}
p.registered <- predict(wls.registered, se.fit = TRUE, validate_d)
error.registered <- (p.registered$fit- validate_d$registered)
RMSE_validation.registered <- sqrt(mean(error.registered^2))
RMSEWLS.registered <- sqrt(mean(resid(wls.registered)^2))

```

```{r}
RMSE_validation.registered
RMSEWLS.registered
```

Relative mean square error
```{r}
mean((error.casual)^2) / mean((validate_d$casual)^2)
mean((error.registered)^2) / mean((validate_d$registered)^2)
```

Our model predicts the bike data in 2012 with mean error of 23 percent and 16 percent within the true value of casual and registered counts respectively. However, our model have twice as large of square root of mean square error with the validation data set than with the training data set. 

```{r}
validate_d <- validate_d %>% mutate(prediction_registered = predict(wls.registered, validate_d)) %>% mutate(prediction_casual = predict(wls.casual, validate_d))

```


```{r}
ggplot(validate_d, aes(x = casual, y = prediction_casual)) + geom_point() +
geom_abline(intercept = 0, slope = 1) +
ggtitle("Validation Casual vs Prediction")
```
```{r}
ggplot(validate_d, aes(x = registered, y = prediction_registered)) + geom_point() +
geom_abline(intercept = 0, slope = 1) +
ggtitle("Validation Registered vs Prediction")
```

```{r}
ggplot(data = validate_d, aes(x = instant)) +
geom_line(aes(y = casual, color = "GroundCO")) +
geom_line(aes(y = prediction_casual, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("registered","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation") 
```

```{r}
ggplot(data = validate_d, aes(x = instant)) +
geom_line(aes(y = registered, color = "GroundCO")) +
geom_line(aes(y = prediction_registered, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("casual","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation")
```

