---
title: "Final"
output: github_document
---

# Introduction

- Interest from a business perspective: helps bike rental businesses meet demands
- City planning perspective: helps cities to adapt to the change of number of bikers to enforce better traffic laws
- A way to sense mobility in the city 

# Backgrounds

The data is a two-year historical in corresponding to years 2011 and 2012 from Capital Bikeshare system, Washington D.C. containing the following datas: 
weathersit:
1: Clear, Few clouds, Partly cloudy,
2: Mist and Cloudy, Mist and Broken clouds, Mist and Few clouds, Mist
3: Light Snow, Light Rain and Thunderstorm and Scattered clouds, Light Rain an dScattered clouds
4: Heavy Rain and Ice Pallets and Thunderstorm and Mist, Snow and Fog
instant: record index

dteday: date

season: season (1:spring, 2:summer, 3:fall, 4:winter)

yr: year (0: 2011, 1:2012)

mnth: month ( 1 to 12)

holiday: weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)

weekday: day of the week

workingday: if day is neither weekend nor holiday is 1, otherwise is 0.

temp: Normalized temperature in Celsius. The values are divided to 41 (max)

atemp: Normalized feeling temperature in Celsius. The values are divided to 50 (max)

hum: Normalized humidity. The values are divided to 100 (max)

windspeed: Normalized wind speed. The values are divided to 67 (max)

casual: count of casual users

registered: count of registered users

cnt: count of total rental bikes including both casual and registered

Our goal is to use data in 2011 to predict bike rential behaviour in 2012.

# Preprocessing

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(GGally)
library(caret)
library(kableExtra)
library(nlme)
library(cowplot)
library(knitr)
```

```{r read file}
bikedata <- read.csv("day.csv",header=T)
names(bikedata)
```

```{r transform data}
#Transform temp, atemp, windspeed, and humidity to actual values
bikedata <- 
  bikedata %>% mutate(actual.temp = temp*41) %>% 
  mutate(actual.atemp = atemp*50) %>%
  mutate(actual.windspeed = windspeed*67) %>%
  mutate(actual.hum = hum*100)

#Combining summer, fall, and spring, winter
bikedata <- bikedata %>% mutate(season.2 = if_else(season == 2|season==3|season==4,0,if_else(season ==1, 1,NA_real_) ))

#process factor data
bikedata$season <- factor(format(bikedata$season, format="%A"),
                   levels = c("1", "2","3","4") , 
                   labels = c("Spring","Summer","Fall","Winter"))

bikedata$spring <- factor(format(bikedata$season.2, format="%A"),
                   levels = c("0","1") , 
                   labels = c("Not Spring","Spring"))

bikedata$holiday <-factor(format(bikedata$holiday, format="%A"),
                          levels = c("0", "1") , 
                          labels = c("Not Holiday","Holiday"))

bikedata$weathersit <- factor(format(bikedata$weathersit, format="%A"),
                       levels = c("1", "2","3","4") , 
                       labels = c("Good:Clear/Sunny","Moderate:Cloudy/Mist","Bad: Rain/Snow/Fog","Worse: Heavy Rain/Snow/Fog"))

bikedata$workingday <- factor(format(bikedata$workingday, format = "%A"), 
                              levels = c("0", "1"),
                              labels = c("Not WorkingDay", "WorkingDay"))

bikedata$yr <- factor(format(bikedata$yr, format="%A"),
                          levels = c("0", "1") , labels = c("2011","2012"))

bikedata <- bikedata %>% mutate(weekend = if_else(weekday == 0|weekday==6,0,if_else(weekday ==1|weekday==2|weekday==3|weekday==4|weekday==5,1,NA_real_) ))

bikedata$weekend <- factor(format(bikedata$weekend, format = "%A"),
                           levels = c(0,1),
                           labels = c("Weekend", "Weekday"))

bikedata$mnth <- as.factor(bikedata$mnth)



#Generate days from start date values
start = "2011-01-01"
bikedata$date_diff <- as.Date(as.character(bikedata$dteday), format="%Y-%m-%d")-
                  as.Date(start, format="%Y-%m-%d")



```


```{r seperate data}
#Seperate training and validating datas base on year

training_d = bikedata %>%  filter(yr == "2011")
set.seed(42)
#partitiontraining <- createDataPartition(y = train$cnt, p = 0.8, list = F)
#training_d <- train[partitiontraining, ]
#test_d <- train[-partitiontraining, ]
validate_d <- bikedata %>% filter(yr == "2012")
```

```{r}
training.workingday = filter(training_d, workingday == "WorkingDay")
training.nworkingday = filter(training_d, workingday == "Not WorkingDay")
validate.workingday = filter(validate_d, workingday == "WorkingDay")
validate.nworkingday = filter(validate_d, workingday == "Not WorkingDay")
```

# Season

```{r}
plot1<- ggplot(training_d,aes(x=season,y=registered))+geom_col()
plot2<- ggplot(training_d,aes(x=season,y=casual ))+geom_col()
plot_grid(plot1, plot2, labels = c("registered", "casual"))
```
```{r}
plot1<- ggplot(training.nworkingday,aes(x=season,y=registered))+geom_col()
plot2<- ggplot(training.nworkingday,aes(x=season,y=casual ))+geom_col()
plot_grid(plot1, plot2, labels = c("registered", "casual"))
```

```{r}
plot1<- ggplot(training.workingday,aes(x=season,y=registered))+geom_col()
plot2<- ggplot(training.workingday,aes(x=season,y=casual ))+geom_col()
plot_grid(plot1, plot2, labels = c("registered", "casual"))
```



The graphs show that for both casual and registered bikers, there are the most rental counts during autumn season and the least during the spring season. However, for registered, there are about the same amount of count during summer and winter while for casual there are significantly less counts during winter than during summer. Therefore we think that we should fit different models for registered and casual.

```{r}
ggplot(training_d,aes(x=season,y=actual.temp))+geom_boxplot()
```

Temperature and seasons are strongly correlated. Spring has the lowest temperature while fall has the highest temperature.





# Holiday, Weekday, Workingday

```{r}
plot1<- ggplot(training_d,aes(x=weekday,y=registered))+geom_col()
plot2 <- ggplot(training_d,aes(x=weekday,y=casual))+geom_col()
plot_grid(plot1, plot2, labels = c("registered", "casual"))
```

Casual rental counts are higher on weekends compared to on weekdays while registered rental counts are lower on weekends than on weekdays. 


```{r}
plot1 <- ggplot(data = training_d, aes(x=date_diff, y = registered)) +  geom_col(aes(fill = workingday))
plot2 <- ggplot(data = training_d, aes(x=date_diff, y = casual)) +  geom_col(aes(fill = workingday))
plot1
plot2
```
There are more rental counts on not working days than on working days for casual bikers while there are more rental registered rental counts on working days than on not workingdays. There are also less rental counts for both registered and casual in the beginning of the year, then we see an increase of bikers during the summer and fall seasons, then a decrease during the end of the year. We suspect that this trend is due to temperature and other weather conditions. 

```{r}
plot1 <- ggplot(data = training_d, aes(x=date_diff, y = registered)) +  geom_col(aes(fill = weathersit))
plot2 <- ggplot(data = training_d, aes(x=date_diff, y = casual)) +  geom_col(aes(fill = weathersit))
plot1
plot2
```
```{r}
plot1 <- ggplot(data = training.nworkingday, aes(x=date_diff, y = registered)) +  geom_col(aes(fill = season))
plot2 <- ggplot(data = training.nworkingday, aes(x=date_diff, y = casual)) +  geom_col(aes(fill = season))
plot1
plot2
```

```{r}
plot1 <- ggplot(data = training.workingday, aes(x=date_diff, y = registered)) +  geom_col(aes(fill = season))
plot2 <- ggplot(data = training.workingday, aes(x=date_diff, y = casual)) +  geom_col(aes(fill = season))
plot1
plot2
```
More bikers tend to bike on days with good and moderate weather conditions than on bad weather conditions. 


```{r}
data <- data.frame(training_d$registered, training_d$actual.hum, training_d$actual.temp, training_d$actual.windspeed, training_d$actual.atemp)

data = data%>% rename( registered = training_d.registered, humidity= training_d.actual.hum,  temperature= training_d.actual.temp, windspeed = training_d.actual.windspeed, feeltemp = training_d.actual.atemp)

plot1 <- ggpairs(data, lower = list(continuous = wrap("points", alpha = 0.3, size= 0.7)))

data <- data.frame(training_d$casual, training_d$actual.hum, training_d$actual.temp, training_d$actual.windspeed, training_d$actual.atemp)

data = data%>% rename( casual = training_d.casual, humidity= training_d.actual.hum,  temperature= training_d.actual.temp, windspeed = training_d.actual.windspeed, feeltemp = training_d.actual.atemp)

plot2 <- ggpairs(data, lower = list(continuous = wrap("points", alpha = 0.3, size= 0.7)))

plot1
plot2
```

The graphs suggest that for both registered and casual bikers, there is a high correlation between temperature, windspeed and rental counts. There is strong correlation between temperature and feel temperature, so we decided to omit feel temperature to avoid collinearity. 


```{r}
m.quadls_casual <- lm(training_d$casual ~ training_d$actual.temp + I(training_d$actual.temp^2))
m.quadls_registered <- lm(training_d$registered ~ training_d$actual.temp + I(training_d$actual.temp^2))

ggplot(training_d, aes(x = actual.temp)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) +   geom_line(data = fortify(m.quadls_casual), aes(x = training_d$actual.temp, y = .fitted), color = "red") + geom_line(data = fortify(m.quadls_registered), aes(x = training_d$actual.temp, y = .fitted), color = "blue") + labs(title = "Scatter plot with fitted models", x = "actual temperature", y = 'rental counts')
```



```{r}
m.casual.workingday <- lm(training.workingday$casual ~ training.workingday$actual.temp)
                        
m.quadls_casual.nworkingday <- lm(training.nworkingday$casual ~ training.nworkingday$actual.temp + I(training.nworkingday$actual.temp^2))

m.registered.nworkingday <- lm(training.nworkingday$registered ~ training.nworkingday$actual.temp + I(training.nworkingday$actual.temp^2))
ggplot(training_d, aes(x = actual.temp)) + geom_point(aes(y = casual,  color = workingday), shape = 1) + geom_line(data = fortify(m.casual.workingday), aes(x = training.workingday$actual.temp, y = .fitted), color = "blue") +  geom_line(data = fortify(m.quadls_casual.nworkingday), aes(x = training.nworkingday$actual.temp, y = .fitted), color = "red") + labs(title = "Scatter plot of casual counts on weekdays and weekends with fitted model")
```
```{r}
m.registered.workingday <- lm(training.workingday$registered ~ training.workingday$actual.temp + I(training.workingday$actual.temp^2))
                        
m.registered.nworkingday <- lm(training.nworkingday$registered ~ training.nworkingday$actual.temp + I(training.nworkingday$actual.temp^2))
ggplot(training_d, aes(x = actual.temp)) + geom_point(aes(y = registered,  color = workingday), shape = 1) + geom_line(data = fortify(m.registered.workingday), aes(x = training.workingday$actual.temp, y = .fitted), color = "blue") +  geom_line(data = fortify(m.registered.nworkingday), aes(x = training.nworkingday$actual.temp, y = .fitted), color = "red") + labs(title = "Scatter plot of registered counts on weekdays and weekends with fitted model")
```

# Wind speed and rental counts

```{r}
m.lin_casual <- lm(training_d$casual ~ training_d$actual.windspeed)
m.lin_registered <- lm(training_d$registered ~ training_d$actual.windspeed)

ggplot(training_d, aes(x = actual.windspeed)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) +   geom_line(data = fortify(m.lin_casual), aes(x = training_d$actual.windspeed, y = .fitted), color = "red") + geom_line(data = fortify(m.lin_registered), aes(x = training_d$actual.windspeed, y = .fitted), color = "blue") + labs(title = "Scatter plot with fitted models", x = 'wind speed', y = 'rental counts')
```


# Humidity

```{r}
m.lin_casual <- lm(training_d$casual ~ training_d$actual.hum) 
m.lin_registered <- lm(training_d$registered ~ training_d$actual.hum )

ggplot(training_d, aes(x = actual.hum)) + geom_point(aes(y = registered,  color = "registered"), shape = 1) + geom_point(aes(y = casual,  color = "casual"), shape = 1) +   geom_line(data = fortify(m.lin_casual), aes(x = training_d$actual.hum, y = .fitted), color = "red") + geom_line(data = fortify(m.lin_registered), aes(x = training_d$actual.hum, y = .fitted), color = "blue") + labs(title = "Scatter plot with fitted models", x = 'humidity', y = 'rental counts')
```
```{r}
m.hum_casual.workingday <- lm(training.workingday$casual ~ training.workingday$actual.hum)
                        
m.hum_casual.nworkingday <- lm(training.nworkingday$casual ~ training.nworkingday$actual.hum)

ggplot(training_d, aes(x = actual.hum)) + geom_point(aes(y = casual,  color = workingday), shape = 1) + geom_line(data = fortify(m.hum_casual.workingday), aes(x = training.workingday$actual.hum, y = .fitted), color = "blue") +  geom_line(data = fortify(m.hum_casual.nworkingday), aes(x = training.nworkingday$actual.hum, y = .fitted), color = "red") + labs(title = "Scatter plot of casual counts on weekdays and weekends with fitted model")
```
```{r}
m.hum_registered.workingday <- lm(training.workingday$registered ~ training.workingday$actual.hum)
                        
m.hum_registered.nworkingday <- lm(training.nworkingday$registered ~ training.nworkingday$actual.hum)

ggplot(training_d, aes(x = actual.hum)) + geom_point(aes(y = registered,  color = workingday), shape = 1) + geom_line(data = fortify(m.hum_registered.workingday), aes(x = training.workingday$actual.hum, y = .fitted), color = "blue") +  geom_line(data = fortify(m.hum_registered.nworkingday), aes(x = training.nworkingday$actual.hum, y = .fitted), color = "red") + labs(title = "Scatter plot of registered counts on weekdays and weekends with fitted model")
```



# Model
```{r}
model.casual.workingday <- lm(log(casual) ~ actual.temp + I(actual.temp^2) + weathersit +actual.windspeed + season, data = training.workingday )

model.registered.workingday <- lm(log(registered) ~   actual.temp + I(actual.temp^2) + weathersit + date_diff , data = training.workingday)
```


```{r}
summary(model.casual.workingday)
```

```{r}
summary(model.registered.workingday)
```

```{r}
model.casual.nworkingday <- lm(log(casual) ~ actual.temp +I(actual.temp^2) + weathersit + actual.windspeed + season, data = training.nworkingday)

model.registered.nworkingday <- lm(log(registered) ~ actual.temp  + weathersit + actual.windspeed, data = training.nworkingday)
```

```{r}
summary(model.casual.nworkingday)
```

```{r}
summary(model.registered.nworkingday)
```



All of the p values on the coefficients of the regressors are less than 0.005. Therefore we are confident that all the regressors have an effect on the rental counts individually. Furthermore, the p value of the F-statistic is less than 0.005. Therefore we are very confident that all the regressors are jointly significant. The R^2 value is arount 0.7, so the models explain around 70 percent of the variation in rental counts. (explain more in paper).

# Model diagnosis

```{r}
StanRes.casual.workingday <- rstandard(model.casual.workingday)
StanRes.registered.workingday <- rstandard(model.registered.workingday)
StanRes.casual.nworkingday <- rstandard(model.casual.nworkingday)
StanRes.registered.nworkingday <- rstandard(model.registered.nworkingday)
```

```{r}
ggplot() +
geom_point(data=training.workingday, aes(x=I(log(casual)), y=StanRes.casual.workingday, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for casual bikers on workingdays")
```

```{r}
ggplot() +
geom_point(data=training.workingday, aes(x=I(log(registered)), y=StanRes.registered.workingday, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for registered bikers on workingdays")
```

```{r}
Fitted_casual.workingday = fitted(model.casual.workingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_casual.workingday, y=StanRes.casual.workingday, color = "MLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals MLS Plot (Fitted) for casual bikers on workingdays") 
```

```{r}
Fitted_registered.workingday = fitted(model.registered.workingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_registered.workingday, y=StanRes.registered.workingday, color = "MLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals MLS Plot (Fitted) for registered bikers on workingdays") 
```


```{r}
ggplot() +
geom_point(data=training.nworkingday, aes(x=log(casual), y=StanRes.casual.nworkingday, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for casual bikers on non-workingdays")
```

```{r}
ggplot() +
geom_point(data=training.nworkingday, aes(x=log(registered), y=StanRes.registered.nworkingday, color = "MLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS"), values = c("blue")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for registered bikers on non-workingdays")
```

```{r}
Fitted_casual.nworkingday = fitted(model.casual.nworkingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_casual.nworkingday, y=StanRes.casual.nworkingday, color = "MLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals MLS Plot (Fitted) for casual bikers on non-workingdays") 
```

```{r}
Fitted_registered.nworkingday = fitted(model.registered.nworkingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_registered.nworkingday, y=StanRes.registered.nworkingday, color = "MLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals MLS Plot (Fitted) for registered bikers on workingdays") 
```


```{r}
p <- ggplot(data.frame(StanRes.casual.workingday), aes(sample = StanRes.casual.workingday)) +
ggtitle("QQ MLS Plot for casual bikers on workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanRes.registered.workingday), aes(sample = StanRes.registered.workingday)) +
ggtitle("QQ MLS Plot for registered bikers on workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanRes.casual.nworkingday), aes(sample = StanRes.casual.nworkingday)) +
ggtitle("QQ MLS Plot for casual bikers on non-workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanRes.registered.nworkingday), aes(sample = StanRes.registered.nworkingday)) +
ggtitle("QQ MLS Plot for registered bikers on non-workingdays")
p + stat_qq() + stat_qq_line()
```



The fitted residual plot and the residual plot suggest that there are extreme outliers in the casual model and that the residual for both models are not evenly distributed around 0, therefore suggesting that there exists heterogeneity in the models. 

The QQ plots show a line that is roughly straight, therefore we conclude that the data of registered bikers come from a normally distributed sample. We can also conclude the same for casual bikers, however, there exists some data points that do not come from a normal distribution as indicated by the few datapoints that deviate significantly from the straight line.


```{r}

ggplot(data = data.frame(StanRes.casual.workingday), aes(x = StanRes.casual.workingday)) + geom_histogram(bins = 50)


```

```{r}
p2 <- ggplot(data = data.frame(StanRes.casual.nworkingday), aes(x = StanRes.casual.nworkingday)) + geom_histogram(bins = 50)

p2
```

```{r}
p3 <- ggplot(data = data.frame(StanRes.registered.workingday), aes(x = StanRes.registered.workingday)) + geom_histogram(bins = 50)

p3
```

```{r}
p4 <- ggplot(data = data.frame(StanRes.registered.nworkingday), aes(x = StanRes.registered.nworkingday)) + geom_histogram(bins = 50)

p4
```

```{r}
acf(StanRes.registered.nworkingday, main="ACF of standardised residuals", 200)
```

```{r}
acf(StanRes.registered.workingday, main="ACF of standardised residuals", 200)
```


```{r}
acf(StanRes.casual.nworkingday, main="ACF of standardised residuals", 200)
```

```{r}
acf(StanRes.casual.workingday, main="ACF of standardised residuals", 200)
```
Therefore using a gls With corrAR1 to correct correlations between y values in different periods. 



# model 2
```{r}
m.gls.casual.workingday <- gls(log(casual) ~ actual.windspeed + actual.temp +I(actual.temp^2) + weathersit + season, data = training.workingday,
correlation=corAR1(form=~instant), method="ML")

summary(m.gls.casual.workingday)
```

```{r}
m.gls.registered.workingday <- gls(log(registered) ~  actual.temp + I(actual.temp^2)+actual.windspeed + weathersit+ date_diff + season, data = training.workingday,
correlation=corAR1(form=~instant), method="ML")

summary(m.gls.registered.workingday)

```

```{r}
m.gls.casual.nworkingday <- gls(log(casual) ~ actual.windspeed + actual.temp +I(actual.temp^2) + weathersit+season , data = training.nworkingday,
correlation=corAR1(form=~instant), method="ML")

summary(m.gls.casual.nworkingday)
```


```{r}
m.gls.registered.nworkingday <- gls(log(registered) ~ actual.temp +  actual.windspeed + weathersit , data = training.nworkingday, correlation=corAR1(form=~instant), method="ML")

summary(m.gls.registered.nworkingday)
```

# Model2 diagnosis
```{r}
StanResGLS.casual.nworkingday <- residuals(m.gls.casual.nworkingday,"pearson")
StanResGLS.casual.workingday <- residuals(m.gls.casual.workingday,"pearson")
StanResGLS.registered.nworkingday <- residuals(m.gls.registered.nworkingday,"pearson")
StanResGLS.registered.workingday <- residuals(m.gls.registered.workingday,"pearson")
```

```{r}
ggplot(data=training.workingday, aes(x=log(casual))) +
geom_point(aes(y=StanRes.casual.workingday, color = "MLS"), size = 1) + 
geom_point(aes(y=StanResGLS.casual.workingday, color = "GLS"), size = 1) + 
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS", "GLS"), values = c("blue", "red")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for casual bikers on workingdays")
```

```{r}
ggplot(data=training.workingday, aes(x=log(registered))) +
geom_point(aes(y=StanRes.registered.workingday, color = "MLS"), size = 1) +
  geom_point(aes(y=StanResGLS.registered.workingday, color = "GLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
scale_color_manual(name = element_blank(), labels = c("MLS", "GLS"), values = c("blue", "red")) +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for registered bikers on workingdays")
```

```{r}
FittedGLS_casual.workingday = fitted(m.gls.casual.workingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_casual.workingday, y=StanRes.casual.workingday, color = "MLS"), size = 1) +
geom_point(aes(x=FittedGLS_casual.workingday, y=StanResGLS.casual.workingday, color = "GLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals Plot (Fitted) for casual bikers on workingdays") 
```

```{r}
FittedGLS_registered.workingday = fitted(model.registered.workingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_registered.workingday, y=StanRes.registered.workingday, color = "MLS"), size = 1) +
geom_point(aes(x=FittedGLS_registered.workingday, y=StanResGLS.registered.workingday, color = "GLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals Plot (Fitted) for registered bikers on workingdays") 
```


```{r}
ggplot(data=training.nworkingday, aes(x=log(casual))) +
geom_point(aes(y=StanRes.casual.nworkingday, color = "MLS"), size = 1) +
geom_point(aes(y=StanResGLS.casual.nworkingday, color = "GLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for casual bikers onnon-workingdays")
```

```{r}
ggplot(data=training.nworkingday, aes(x=log(registered))) +
geom_point(aes(y=StanRes.registered.nworkingday, color = "MLS"), size = 1) +
geom_point(aes(y=StanResGLS.registered.nworkingday, color = "GLS"), size = 1) +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') +
labs(y = "Standarized Residual") + ggtitle("Standarized Residuals MLS Plot for registered bikers on non-workingdays")
```

```{r}
FittedGLS_casual.nworkingday = fitted(model.casual.nworkingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_casual.nworkingday, y=StanRes.casual.nworkingday, color = "MLS"), size = 1) +
geom_point(aes(x=FittedGLS_casual.nworkingday, y=StanResGLS.casual.nworkingday, color = "GLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals Plot (Fitted) for casual bikers on non-workingdays") 
```

```{r}
FittedGLS_registered.nworkingday = fitted(model.registered.nworkingday)

ggplot() +
geom_hline(yintercept=2,color='blue') + geom_hline(yintercept=-2, color='blue') + geom_point(aes(x=Fitted_registered.nworkingday, y=StanRes.registered.nworkingday, color = "MLS"), size = 1) +
geom_point(aes(x=FittedGLS_registered.nworkingday, y=StanResGLS.registered.nworkingday, color = "GLS"), size = 1) +
labs(y = "Standarized Residual") + labs(x = "Fitted value") +
ggtitle("Standarized Residuals Plot (Fitted) for registered bikers on workingdays") 
```


```{r}
p <- ggplot(data.frame(StanResGLS.casual.workingday), aes(sample = StanResGLS.casual.workingday)) +
ggtitle("QQ MLS Plot for casual bikers on workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanResGLS.registered.workingday), aes(sample = StanResGLS.registered.workingday)) +
ggtitle("QQ MLS Plot for registered bikers on workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanResGLS.casual.nworkingday), aes(sample = StanResGLS.casual.nworkingday)) +
ggtitle("QQ MLS Plot for casual bikers on non-workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p <- ggplot(data.frame(StanResGLS.registered.nworkingday), aes(sample = StanResGLS.registered.nworkingday)) +
ggtitle("QQ MLS Plot for registered bikers on non-workingdays")
p + stat_qq() + stat_qq_line()
```

```{r}
p1 <- ggplot(data = data.frame(StanResGLS.casual.workingday), aes(x = StanResGLS.casual.workingday)) + geom_histogram(bins = 50)

p1
```

```{r}
p2 <- ggplot(data = data.frame(StanResGLS.casual.nworkingday), aes(x = StanResGLS.casual.nworkingday)) + geom_histogram(bins = 50)

p2
```

```{r}
p3 <- ggplot(data = data.frame(StanResGLS.registered.workingday), aes(x = StanResGLS.registered.workingday)) + geom_histogram(bins = 50)

p3
```

```{r}
p4 <- ggplot(data = data.frame(StanResGLS.registered.nworkingday), aes(x = StanResGLS.registered.nworkingday)) + geom_histogram(bins = 50)

p4
```


# Validation with model 2

```{r}

```

```{r}
p.casual.workingday <- exp(predict(model.casual.workingday,validate.workingday))
error.casual.workingday <- ((p.casual.workingday)- validate.workingday$casual)
RMSE_validation.caual.workingday <- log(mean(error.casual.workingday^2))
pt.casual.workingday <- exp(predict(model.casual.workingday,training.workingday))
errort.casual.workingday <- ((pt.casual.workingday)- training.workingday$casual)
RMSE.casual.workingday <- log(mean(errort.casual.workingday)^2)

p.casual.nworkingday <- exp(predict(model.casual.nworkingday, validate.nworkingday))
error.casual.nworkingday <- (p.casual.nworkingday- validate.nworkingday$casual)
RMSE_validation.caual.nworkingday <- log(mean(error.casual.nworkingday^2))
pt.casual.nworkingday <- exp(predict(model.casual.nworkingday, training.nworkingday))
errort.casual.nworkingday <- (pt.casual.nworkingday- training.nworkingday$casual)
RMSE.casual.nworkingday <- log(mean(errort.casual.nworkingday)^2)

```

Square root mean square error for validation data set
```{r}
RMSE_validation.caual.workingday
RMSE_validation.caual.nworkingday
```
square root mean square error for training data set
```{r}
RMSE.casual.workingday
RMSE.casual.nworkingday
```

```{r}
p.registered.workingday <- exp(predict(model.registered.workingday,validate.workingday))
error.registered.workingday <- ((p.registered.workingday)- validate.workingday$registered)
RMSE_validation.registered.workingday <- log(mean(error.registered.workingday^2))
pt.registered.workingday <- exp(predict(model.registered.workingday,training.workingday))
errort.registered.workingday <- ((pt.registered.workingday)- training.workingday$registered)
RMSE.registered.workingday <- log(mean(errort.registered.workingday)^2)

p.registered.nworkingday <- exp(predict(m.gls.registered.nworkingday, validate.nworkingday))
error.registered.nworkingday <- (p.registered.nworkingday- validate.nworkingday$registered)
RMSE_validation.registered.nworkingday <- log(mean(error.registered.nworkingday^2))
pt.registered.nworkingday <- exp(predict(model.registered.nworkingday, training.nworkingday))
errort.registered.nworkingday <- (pt.registered.nworkingday- training.nworkingday$registered)
RMSE.registered.nworkingday <- log(mean(errort.registered.nworkingday)^2)


```

Square root mean square error for validation data set

```{r}
RMSE_validation.registered.workingday
RMSE_validation.registered.nworkingday
```
square root mean square error for training data set

```{r}
RMSE.registered.workingday
RMSE.registered.nworkingday
```


Relative mean square error
```{r}
mean((error.casual.workingday)^2) / mean((validate.workingday$casual)^2)
mean((error.casual.nworkingday)^2) / mean((validate.nworkingday$casual)^2)
mean((error.registered.workingday)^2) / mean((validate.workingday$registered)^2)
mean((error.registered.nworkingday)^2) / mean((validate.nworkingday$registered)^2)


```

Our model predicts the bike data in 2012 with mean error of 23 percent and 16 percent within the true value of casual and registered counts respectively. However, our model have twice as large of square root of mean square error with the validation data set than with the training data set. 

```{r}
validate.workingday <- validate.workingday %>% mutate(prediction_registered.workingday =  exp(predict(model.registered.workingday, validate.workingday))) %>% mutate(prediction_casual.workingday = exp(predict(model.casual.workingday, validate.workingday)))

validate.nworkingday <- validate.nworkingday %>% mutate(prediction_registered.nworkingday = exp(predict(model.registered.nworkingday, validate.nworkingday))) %>% mutate(prediction_casual.nworkingday = exp(predict(model.casual.nworkingday, validate.nworkingday)))
```


```{r}
ggplot(validate.workingday, aes(x = casual, y = prediction_casual.workingday)) + geom_point() +
geom_abline(intercept = 0, slope = 1) +
ggtitle("Validation Casual vs Prediction on workingdays") +
ylab("prediction")
```

```{r}
ggplot(validate.nworkingday, aes(x = casual, y = prediction_casual.nworkingday)) + geom_point() +
geom_abline(intercept = 0, slope = 1) +
ggtitle("Validation Casual vs Prediction on non-workingdays")+
ylab("prediction")
```

```{r}
ggplot(validate.workingday, aes(x = registered, y = prediction_registered.workingday)) + geom_point() +
geom_abline(intercept = 0, slope = 1) +
ggtitle("Validation Registered vs Prediction on workingdays")+
ylab("prediction")
```

```{r}
ggplot(validate.nworkingday, aes(x = registered, y = prediction_registered.nworkingday)) + geom_point() +
geom_abline(intercept = 0, slope = 1) +
ggtitle("Validation Registered vs Prediction on non-workingdays")+
ylab("prediction")
```

```{r}
ggplot(data = validate.workingday, aes(x = instant)) +
geom_line(aes(y = casual, color = "casual")) +
geom_line(aes(y = prediction_casual.workingday, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("casual","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation of casual bikers on workingdays") 
```


```{r}
ggplot(data = validate.nworkingday, aes(x = instant)) +
geom_line(aes(y = casual, color = "casual")) +
geom_line(aes(y = prediction_casual.nworkingday, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("casual","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation of casual bikers on non-workingdays") 
```

```{r}
ggplot(data = validate.workingday, aes(x = instant)) +
geom_line(aes(y = registered, color = "groundtruth")) +
geom_line(aes(y = prediction_registered.workingday, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("groundtruth","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation of registered bikers on workingdays")
```

```{r}
ggplot(data = validate.nworkingday, aes(x = instant)) +
geom_line(aes(y = registered, color = "groundtruth")) +
geom_line(aes(y = prediction_registered.nworkingday, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("groundtruth","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation of registered bikers on non-workingdays")
```


```{r}
validate.nworkingday<- validate.nworkingday %>% mutate(pred.total = prediction_registered.nworkingday+prediction_casual.nworkingday)

validate.workingday<- validate.workingday %>% mutate(pred.total = prediction_registered.workingday+prediction_casual.workingday)

temp1<- subset(validate.nworkingday, select = c(instant,pred.total, cnt))
temp2<- subset(validate.workingday, select = c(instant,pred.total, cnt))
total<- rbind(temp1, temp2)

```


```{r}
ggplot(data = total, aes(x = instant)) +
geom_line(aes(y = cnt, color = "GroundTruth")) +
geom_line(aes(y = pred.total, color="Prediction"), linetype="twodash") +
scale_color_manual(name = element_blank(), labels = c("GroundTruth","Prediction"),
values = c("darkred", "steelblue")) + labs(y = "") +
ggtitle("Validation of total rental counts")
```
```{r}
error.casual.nworkingday <- (total$pred.total- total$cnt)
mean((error.casual.workingday)^2) / mean((validate.workingday$casual)^2)
```


